---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
###navigate to the appropriate directory
```{r}
setwd("/Users/timbo/coursera/repro/Project1")
```

###download file
```{r}
if (!file.exists("./repdata-data-activity.zip")){
        download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",
                      "repdata-data-activity.zip",method="curl")
        
}
```

###unzip file
```{r}
unzip("repdata-data-activity.zip")
```

###read file into R
```{r}
activity <- read.csv("activity.csv")
```

###load libraries
```{r}
library(ggplot2) #for plotting
library(Rmisc) #for the multiplot function
```

###convert the columns into date and time objects
```{r}
activity$date <- as.Date(as.character(activity$date))
activity$interval <- sprintf("%04d",activity$interval)
activity$interval <- as.POSIXlt(activity$interval,format = "%H%M")
activity$interval <- format(activity$interval,"%H:%M")
```

###create data.frame with NA values removed
```{r}
no_na_act <- activity[which(complete.cases(activity)),]
```

## What is mean total number of steps taken per day?
### Calculate total number of steps for each day, na removed
```{r}
daily_sums <- plyr::ddply(no_na_act,"date",
                          plyr::summarise,sum = sum(steps))
names(daily_sums) <- c("date","steps")
```

### Plot histogram
```{r}
ggplot(daily_sums,aes(steps))+
        xlab("Steps Per Day")+geom_histogram()
```

###Calculate mean and median per day
```{r}
summary(daily_sums$steps)
```

## What is the average daily activity pattern?
### get mean for each interval
```{r}
interval_mean <- plyr::ddply(no_na_act,"interval",
                             plyr::summarise,mean = mean(steps))
```

###plot time series
```{r}
ggplot(interval_mean,aes(interval,mean,group=1))+
        scale_x_discrete(breaks=c("03:00","06:00","09:00","12:00",
                                  "15:00","18:00","21:00"))+
        geom_line()
```

### Find interval with highest mean number of steps
```{r}
interval_mean[which(interval_mean$mean==max(interval_mean$mean)),]
```

## Imputing missing values
### Calculate number of rows with missing data
```{r}
steps_na<-sum(is.na(activity$steps))
date_na<-sum(is.na(activity$date))
interval_na<-(is.na(activity$interval))
sum(c(steps_na,date_na,interval_na))
```

### Replace missing value with mean for that interval-choosing this approach because of the degree of variablility among different intervals
```{r}
activity_filled <- activity # make a copy
```

### Replace rows missing steps by taking mean for a given interval
```{r}
activity_filled$steps <- ifelse(is.na(activity_filled$steps),
                        interval_mean$mean[interval_mean$interval
                        %in% activity_filled$interval],
                        activity_filled$steps)
```

### Daily sums using imputed data
```{r}
daily_sums_filled <- plyr::ddply(activity_filled,"date",
                                 plyr::summarise,sum = sum(steps))
names(daily_sums_filled) <- c("date","steps")
```

### Original histogram - NA removed
```{r}
h1 <- ggplot(daily_sums,aes(steps))+
        xlab("Steps Per Day - NA Removed")+geom_histogram()
```

### New histogram - NA imputed
```{r}
h2 <- ggplot(daily_sums_filled,aes(steps))+
        xlab("Steps Per Day - NA Imputed")+geom_histogram()
```

### Plot histograms side-by-side for comparison
```{r}
multiplot(h1,h2)
```

### Compare median and means
```{r}
summary(daily_sums$steps) #old data
summary(daily_sums_filled$steps) #new data
```

## Are there differences in activity patterns between weekdays and weekends?
### Set up weekday variable
```{r}
activity_filled$day_type <- ifelse(weekdays(activity_filled$date)       
                                   %in% c("Monday","Tuesday","Wednesday",
                                          "Thursday","Friday"),
                                   activity_filled$day_type <- "weekday",
                                   activity_filled$day_type <- "weekend")
```

### Subset data plot
```{r}
weekday_df <- activity_filled[which(activity_filled$day_type=="weekday"),]
weekend_df <- activity_filled[which(activity_filled$day_type=="weekend"),]
```

### Calculate mean for each data set
```{r}
wd_interval_mean <- plyr::ddply(weekday_df,"interval",
                             plyr::summarise,mean = mean(steps))
we_interval_mean <- plyr::ddply(weekend_df,"interval",
                                plyr::summarise,mean=mean(steps))
```

### Create plots
```{r}
wd_ts <- ggplot(wd_interval_mean,aes(interval,mean,group=1))+
        scale_x_discrete(breaks=c("03:00","06:00","09:00","12:00",
                                  "15:00","18:00","21:00"))+
        ylim(0,250)+
        ggtitle("Weekday Step Count by Interval")+
        geom_line()

we_ts <- ggplot(we_interval_mean,aes(interval,mean,group=1))+
        scale_x_discrete(breaks=c("03:00","06:00","09:00","12:00",
                                  "15:00","18:00","21:00"))+
        ylim(0,250)+
        ggtitle("Weekend Step Count by Interval")+
        geom_line()

# display plots
multiplot(wd_ts,we_ts)
```